---
title: "Final Project - Classifying New York Census Tracts"
author: "Sunduss Hamdan and Kaela Girod"
format:
  html:
    embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## 1. Introduction
New York City is one of the most unequal cities in the United States. While some 
neighborhoods are home to some of the wealthiest residents in the world, others 
face persistent poverty, limited job opportunities, and inadequate access to 
resources. Despite this stark divide, policymakers tasked with deciding where to 
direct funding and interventions often rely on a single measure, such as average 
income or the share of residents living below the poverty line, to characterize 
neighborhoods and determine need. The problem with this approach is that it 
flattens the complexity of what makes neighborhoods genuinely different from one 
another. A neighborhood with a high poverty rate but stable housing, good transit, 
and strong employment access is a very different place from one with the same 
poverty rate but high vacancy, poor connectivity, and limited opportunity. Treating 
them the same way leads to poorly targeted policies. This project takes a different 
approach: we use data-driven methods, specifically using three models 
(Principle Component Analysis, K-means clustering, and Gaussian Mixture Modeling) to identify distinct neighborhood typologies 
across New York City's census tracts, asking what distinct socioeconomic typologies 
exist across NYC census tracts, and what combinations of demographic and economic 
characteristics define them.

## 2. Background, Policy Relevance, and Literature
New York City's inequality is well documented. The top three percent of wage earners 
in the city have seen their incomes grow substantially, while low-wage workers have 
faced increasing risk of poverty (Center for New York City Affairs, 2024). This 
inequality is not just economic, it is also deeply racial. The New York City 
Comptroller (2023) reports that the racial wealth gap in New York is among the most 
pronounced of any major American city, with Black and Hispanic residents 
significantly less likely to own homes, hold financial assets, or access 
higher-paying jobs than their white counterparts. These disparities do not play out 
evenly across the city; they are concentrated in particular neighborhoods, making 
the neighborhood itself a critical unit of analysis for anyone interested in 
understanding or addressing inequality in New York City.

Despite this, the tools that policymakers most commonly use to identify and target 
disadvantaged neighborhoods remain blunt. Relying on a single indicator like median 
income or the poverty rate to decide where to direct resources misses the fact that 
neighborhoods can be disadvantaged in very different ways. A neighborhood with high 
poverty but stable housing and good transit access has different needs, and 
different strengths, than one with similar poverty levels but high vacancy rates, 
poor connectivity, and a declining population. Effective place-based policy requires 
a more nuanced understanding of what makes each neighborhood distinct.

The academic literature on neighborhood classification offers a more rigorous 
alternative. Scally et al. (2021) used data-driven classification methods to 
identify distinct typologies among rural census tracts across the United States, 
grouping communities by their combinations of characteristics rather than by any 
single measure. Their central insight, that communities cluster into meaningfully 
different types, and that policymakers should act on that structure, forms the 
foundation of our approach. While their work focused on rural America, the core 
logic translates directly to an urban setting: New York City's neighborhoods are no 
more uniform than rural communities, and a typology-based approach offers the same 
potential to improve how investments are targeted.

```{r}
#| echo: false
#| message: false
#| warning: false
### MAYBE INCLUDE SOME EXPLORATORY AND EXPLANATORY ANALYSIS VISUALS OF KEY VARIABLES OR DESCRIPTIVE STATS ON NEW YORK - SUNDUSS CAN DO THIS ONCE THE DATASET IS UP AND RUNNING 
```

## 3. Data and Variable Selection - EDIT THE BELOW ONCE WE ADD ADDITIONAL VARIABLES TO THE DATASET E.G., SHOOTINGS 
Our primary dataset comes from the 2020–2024 American Community Survey (ACS) 
5-year estimates, which we access via the Census Bureau API at the census tract 
level across all five boroughs of New York City. The ACS 5-year estimates pool 
responses collected over a five-year period, making them the most reliable source 
of small-area data for geographies as granular as census tracts. Following Scally 
et al. (2021), we select variables across five thematic dimensions - economic 
conditions, educational attainment, housing, racial and ethnic composition, and 
mobility - to capture what makes neighborhoods genuinely distinct from one another, 
rather than relying on any single measure like income or poverty alone.

Many ACS variables are provided as raw counts, which reflect tract population size 
as much as neighborhood character. We therefore convert these to shares or rates. 
For example, expressing the number of residents below the poverty line as a 
proportion of all residents for whom poverty status is determined. Two variables, 
median household income and median gross rent, are kept as dollar values since they 
are already directly comparable across tracts. Mean commute time is calculated by 
dividing total travel time (B08013) by the number of workers aged 16 and over who 
commute outside the home (B08012), which is consistent with the Census Bureau's 
own approach to computing this measure.

We considered but decided to exclude two variables. First, the Gini coefficient (B19083), 
which measures income inequality within a tract, was dropped because it captures 
information already covered by our combination of median income and poverty rate. 
Second, age structure variables (B01001) were excluded because including multiple 
age-by-sex breakdown columns would add complexity without adding interpretive value; 
educational attainment serves as a cleaner proxy for the working-age population. 
Both decisions follow a simple guiding principle: where two variables tell 
essentially the same story, we keep the one that is easier to interpret in a 
policy context. [EXPLAIN WHY WE DIDN'T CHOOSE THE ARRESTS VARIABLE AND USED 
SHOOTINGS INSTEAD]

In addition to our primary 2020–2024 dataset, we also extract the 2018–2022 ACS 
5-year estimates for the same tracts. This allows us to compute four supplementary 
change variables in median income, median rent, poverty rate, and unemployment 
rate, that measure how conditions shifted between the two periods. These change 
variables are not used as inputs to our clustering models; instead, they are used 
afterwards to help describe and validate the typologies we identify. One limitation 
worth noting is that because both ACS vintages include data collected between 2020 
and 2022, they share three overlapping years of data. This means the change 
variables capture a narrower window of genuine change than the four-year gap between 
the vintage endpoints might suggest. This is a structural feature of ACS 5-year 
estimates rather than a choice we made, and we flag it as a limitation of our 
approach.

Our [EDIT WITH OUR ACTUAL NO. OF VARIABLES] modelling variables are used as inputs 
to our dimensionality reduction and clustering models. Census tracts with missing values 
on any modelling variable are excluded from the analysis. Missing data arise most 
commonly in tracts that are entirely renter-occupied, which have no median home 
value, and in tracts with very few commuters, which have no reliable commute time 
estimate.

```{r}
#| label: tbl-variables
#| tbl-cap: "Variable Selection and Justification"
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(kableExtra)

variable_table <- tribble(
  ~Dimension,        ~Variable,                  ~Construction,                                          ~Justification,
  "Economic",        "Poverty rate",              "Pop. below poverty line / poverty universe total",     "Core deprivation indicator; directly named in proposal",
  "Economic",        "Unemployment rate",         "Unemployed civilians / civilian labor force",          "Captures labor market exclusion distinct from income poverty",
  "Economic",        "Median household income",   "Level (dollars)",                                      "Primary measure of economic resources; retained as level as it is already comparable across tracts",
  "Education",       "Bachelor's degree share",   "Bachelor's holders / pop. 25+",                       "Captures human capital; strongly associated with neighborhood economic trajectory",
  "Housing",         "Median gross rent",         "Level (dollars)",                                      "Directly named in proposal; key measure of housing cost burden",
  "Housing",         "Median home value",         "Level (dollars)",                                      "Complements rent; distinguishes owner- vs. renter-dominated neighborhoods",
  "Housing",         "Renter-occupied share",     "Renter units / total occupied units",                  "Tenure structure is a fundamental dimension of NYC neighborhood type",
  "Housing",         "Vacancy rate",              "Vacant units / total housing units",                   "Indicator of disinvestment; key differentiator between declining and stable neighborhoods",
  "Race/Ethnicity",  "Share White",               "White alone / total population",                       "Captures racial composition; connected to NYC's documented racial wealth gap",
  "Race/Ethnicity",  "Share Black",               "Black alone / total population",                       "As above",
  "Race/Ethnicity",  "Share Hispanic",            "Hispanic or Latino / total population (B03003)",       "Uses B03003 rather than B02001 as Hispanic identity is an ethnicity not captured by race alone",
  "Race/Ethnicity",  "Share foreign-born",        "Foreign-born / total population",                      "NYC's immigrant geography is a major structural driver of neighborhood type",
  "Mobility",        "Mean commute time",         "Aggregate travel time / commuting workers 16+",        "Directly named in proposal; proxy for transit access and economic integration",
  "Supplementary",   "Change in median income",   "% change, 2022-2024 vintages",                         "Post-hoc cluster validation only; not a modelling input",
  "Supplementary",   "Change in median rent",     "% change, 2022-2024 vintages",                         "As above",
  "Supplementary",   "Change in poverty rate",    "Percentage point change",                              "As above; percentage point change is appropriate for rates",
  "Supplementary",   "Change in unemployment",    "Percentage point change",                              "As above")

variable_table |>
  kbl(
    col.names = c("Dimension", "Variable", "Construction", "Justification"),
    align = c("l", "l", "l", "l")) |>
   kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 13) |>
  column_spec(1, bold = TRUE, width = "10em") |>
  column_spec(2, width = "12em") |>
  column_spec(3, width = "14em") |>
  column_spec(4, width = "20em") |>
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") |>
  pack_rows("Modelling Variables", 1, 13) |>
  pack_rows("Supplementary Variables — Used After Clustering To Validate Typologies, Not As Modelling Inputs", 14, 17)
```

## 4. Methodology
Our analysis uses three methods in sequence: Principal Component Analysis (PCA) 
to reduce the complexity of our data, K-means clustering to identify an initial 
set of neighborhood groups, and Gaussian Mixture Modeling (GMM) to produce a 
more statistically rigorous final classification. Each step is described below.

### 4.1. Step 1: Dimensionality Reduction with PCA
We begin with [INSERT NO. OF FINAL VARIABLES] variables across five thematic dimensions. Many of these 
variables are likely to be correlated with one another. For example, poverty 
rate and unemployment rate tend to move together, as do median income and 
bachelor's degree share. When highly correlated variables are fed directly into 
a clustering algorithm, they effectively get counted twice, overweighting the 
dimension of variation they share and distorting the resulting groups.

To address this, we use Principal Component Analysis (PCA) before clustering. 
PCA takes our [NO. OF VARIABLES] variables and transforms them into a smaller set of new 
variables called principal components. Each component is a weighted combination 
of the original variables, and the components are constructed so that they are 
uncorrelated with one another. The first component captures the largest share 
of overall variation in the data, the second captures the next largest share, 
and so on. Together, a small number of components can often represent most of 
the information in the original data with much less redundancy. Because our variables are measured in very different units (dollars, 
percentages, minutes) we normalize all variables before running PCA by 
subtracting the mean and dividing by the standard deviation so that no single variable dominates due to its scale alone.

To decide how many components to retain, we use a scree plot and inspect the 
cumulative proportion of variance explained by each additional component. We 
retain enough components to explain at least 80% [THIS WILL NEED TO BE CHANGED WITH OUR ACTUAL %] of the total variation in the 
data, following standard practice. The retained components are then passed 
forward as inputs to the clustering stage.

### 4.2. Step 2: K-Means Clustering

We use K-means clustering as our first clustering step. K-means works by 
grouping census tracts into a fixed number of clusters, where each tract is 
assigned to the cluster whose center (called the centroid) it is closest to. 

The algorithm works as follows:
1. A set of K centroids are placed randomly in the data space
2. Each tract is assigned to its nearest centroid based on Euclidean distance
3. The centroids are recalculated as the mean of all tracts assigned to them
4. Steps 2 and 3 repeat until no tracts change cluster

Because the starting positions of the centroids are random, the results of 
K-means are not fully deterministic, running the algorithm twice can produce 
different solutions. We address this by using 100 random starts (`nstart = 100`) 
and retaining the best solution, which is standard practice.

K-means also requires us to specify the number of clusters K in advance. To 
choose K, we use three complementary methods. The first is the total 
within-cluster sum of squares, which measures how tightly packed each cluster 
is, we look for an "elbow" in the plot where adding more clusters stops 
producing meaningful improvements. The second is the average silhouette width, 
which measures how well each tract fits its assigned cluster compared to the 
nearest alternative cluster, with higher values indicating better-defined 
clusters. The third is the gap statistic, which compares the within-cluster 
dispersion to a reference distribution representing data with no real cluster 
structure, with higher values indicating more meaningful clusters. Where these 
three methods agree, we have greater confidence in our chosen K.

We note two important limitations of K-means. First, it assigns every tract to 
exactly one cluster with no measure of uncertainty; a tract that sits on the 
boundary between two neighborhood types is treated the same as one that sits 
squarely in the middle of a cluster. Second, as Milechin et al. (2024) found in 
a closely related study, the statistically optimal number of clusters and the 
most interpretable number are not always the same. We therefore treat K-means 
as an exploratory step and use its results to inform, but not determine, our 
final model.

### 4.3. Step 3: Gaussian Mixture Modeling

Our final and primary clustering method is Gaussian Mixture Modeling (GMM), 
implemented using `library(mclust)`. GMM extends the logic of K-means in two 
important ways that make it better suited to our goals.

First, instead of measuring the distance from each tract to a centroid, GMM 
models each cluster as a multivariate normal distribution, meaning each cluster 
has its own center and its own shape, capturing the spread and correlation 
structure of the tracts within it. Tracts that are close to the center of a 
cluster's distribution are more likely to belong to it; tracts far from the 
center are less likely.

Second, GMM produces soft assignments rather than hard ones. Rather than 
assigning each tract definitively to a single cluster, GMM estimates the 
probability that each tract belongs to each cluster. These probabilities (called 
responsibilities) sum to one for each tract. A tract with a responsibility of 
0.95 for cluster 3 belongs there with high confidence; a tract with 
responsibilities of 0.52 and 0.48 across two clusters is genuinely ambiguous, 
and GMM makes that uncertainty visible rather than hiding it.

To select the number of clusters, we use the Bayesian Information Criterion 
(BIC). BIC works by balancing model fit against model complexity. A model with 
more clusters will always fit the data better, but BIC penalizes the additional 
parameters required, steering us toward solutions that are as simple as possible 
while still capturing meaningful structure in the data. In `library(mclust)`, 
the function automatically tries between 1 and 9 clusters across 14 different 
variance model specifications, and we select the model with the highest BIC 
value reported by the package. One small technical note: the standard statistical 
definition of BIC says lower values are better, but the library(mclust) package 
displays BIC with the opposite sign, meaning we look for the highest value on its 
output plots. Nevertheless, the end result is the same; we are still selecting the model that best balances fit and simplicity.
As with K-means, we do not automatically accept the number of clusters that scores best on BIC. If the statistically preferred solution produces neighborhood types that are too similar to one another to be meaningfully different in a policy context, we will consider a solution with fewer clusters and explain our reasoning for doing so.

## 5. Results

```{r}
#| echo: false
#| message: false
#| warning: false
### INCLUDE VISUALS HERE
```

## 6. Discussion

```{r}
#| echo: false
#| message: false
#| warning: false
### INCLUDE VISUALS HERE
```

## 7. Policy Implications [combine with conclusion if needs be]

## 8. Conclusion

## References
Center for New York City Affairs. "While the Top Three Percent of Wage Earners Get Richer, New York City's Low Wage Workers Risk Greater Poverty." 2024. https://www.centernyc.org/reports-briefs/while-the-top-three-percent-of-wage-earners-get-richer-new-york-citys-low-wage-workers-risk-greater-poverty

New York City Comptroller. "The Racial Wealth Gap in New York." December 6, 2023. https://comptroller.nyc.gov/reports/the-racial-wealth-gap-in-new-york/

Scally, Corianne Payton, Amanda Gold, Yipeng Su, Jorge Morales-Burnett, Eric Burnstein, and Patrick Spauster. "Reenvisioning Rural America: How to Invest in the Strengths and Potential of Rural Communities." 2021. https://reenvisioning-rural-america.urban.org/

U.S. Census Bureau. American Community Survey 5-Year Estimates, 2018–2022 and 2020–2024. Accessed via Census Bureau API using tidycensus in R.

Zewdie, Hiwot Y., Jamaica R. Robinson, Marc A. Adams, Anjum Hajat, Jana A. Hirsch, Brian E. Saelens, and Stephen J. Mooney. "A Tale of Many Neighborhoods: Latent Profile Analysis to Derive a National Neighborhood Typology for the US." Health & Place 86 (March 2024): 103209. https://pubmed.ncbi.nlm.nih.gov/38408408/