---
title: "Classifying New York City Neighborhoods: A Data-Driven Approach to Urban Typologies"
author: "Sunduss Hamdan and Kaela Girod"
format: html
embed-resources: true
execute:
  warning: false
  message: false
editor: visual
editor_options: 
  chunk_output_type: console
---

## 1. Introduction
New York City is one of the most unequal cities in the United States. While some 
neighborhoods are home to some of the wealthiest residents in the world, others 
face persistent poverty, limited job opportunities, and inadequate access to 
resources. Despite this stark divide, policymakers tasked with deciding where to 
direct funding and interventions often rely on a single measure, such as average 
income or the share of residents living below the poverty line, to characterize 
neighborhoods and determine need. The problem with this approach is that it 
flattens the complexity of what makes neighborhoods genuinely different from one 
another. A neighborhood with a high poverty rate but stable housing, good transit, 
and strong employment access is a very different place from one with the same 
poverty rate but high vacancy, poor connectivity, and limited opportunity. Treating 
them the same way leads to poorly targeted policies. This project takes a different 
approach: we use data-driven methods, specifically using two dimension reduction techniques (Generalized Low Rand Models and Principle Component Analysis) along with two clustering techniques (Gaussian Mixture Modeling and K-Means Clustering) to identify distinct neighborhood typologies across New York City's census tracts, asking what combinations of demographic and economic characteristics define them.

## 2. Background, Policy Relevance, and Literature
New York City's inequality is well documented. The top three percent of wage earners 
in the city have seen their incomes grow substantially, while low-wage workers have 
faced increasing risk of poverty (Center for New York City Affairs, 2024). This 
inequality is not just economic, it is also deeply racial. The New York City 
Comptroller (2023) reports that the racial wealth gap in New York is among the most 
pronounced of any major American city, with Black and Hispanic residents 
significantly less likely to own homes, hold financial assets, or access 
higher-paying jobs than their white counterparts. These disparities do not play out 
evenly across the city; they are concentrated in particular neighborhoods, making 
the neighborhood itself a critical unit of analysis for anyone interested in 
understanding or addressing inequality in New York City.

Despite this, the tools that policymakers most commonly use to identify and target 
disadvantaged neighborhoods remain blunt. Relying on a single indicator like median 
income or the poverty rate to decide where to direct resources misses the fact that 
neighborhoods can be disadvantaged in very different ways. A neighborhood with high 
poverty but stable housing and good transit access has different needs, and 
different strengths, than one with similar poverty levels but high vacancy rates, 
poor connectivity, and a declining population. Effective place-based policy requires 
a more nuanced understanding of what makes each neighborhood distinct.

The academic literature on neighborhood classification offers a more rigorous 
alternative. Scally et al. (2021) used data-driven classification methods to 
identify distinct typologies among rural census tracts across the United States, 
grouping communities by their combinations of characteristics rather than by any 
single measure. Their central insight, that communities cluster into meaningfully 
different types, and that policymakers should act on that structure, forms the 
foundation of our approach. While their work focused on rural America, the core 
logic translates directly to an urban setting: New York City's neighborhoods are no 
more uniform than rural communities, and a typology-based approach offers the same 
potential to improve how investments are targeted.

Figure 1 below illustrates the distribution of poverty rates across census tracts within each NYC borough. The Bronx stands out as the most deprived borough, with both the highest median poverty rate and the widest spread across its tracts, indicating persistent and widespread deprivation. Manhattan presents a particularly striking pattern; despite having a relatively low median poverty rate, it contains numerous high-poverty outlier tracts, reflecting the extreme inequality between its wealthiest and most deprived neighborhoods. Even in comparatively affluent boroughs like Queens and Staten Island, individual tracts with very high poverty rates exist. This variation (both between and within boroughs) shows exactly why relying on a single indicator or aggregating data at the borough level is insufficient for effective policy targeting. A more granular, data-driven approach that captures the full complexity of neighborhood conditions is needed, which is what this project sets out to provide.

```{r}
#| echo: false
#| message: false
#| results: false
#| warning: false

library(tidyverse)
library(sf)

nyc_final <- st_read("data/nyc_final.shp")

nyc_data <- nyc_final |>
  st_drop_geometry()

nyc_data <- nyc_data |>
  mutate(borough = case_when(
    contyfp == "005" ~ "Bronx",
    contyfp == "047" ~ "Brooklyn",
    contyfp == "061" ~ "Manhattan",
    contyfp == "081" ~ "Queens",
    contyfp == "085" ~ "Staten Island"))

ggplot(nyc_data |> filter(!is.na(borough)), 
       aes(x = borough, y = pct_pvr, fill = borough)) +
  geom_boxplot() +
  labs(
    title = "The Bronx Has the Highest and Most Unequal Poverty Rates \nAcross NYC Boroughs",
    subtitle = "\nEach box shows the distribution of poverty rates across census tracts within each borough.\nThe middle line is the median tract, the box covers the middle 50%, and dots are outliers.",
    x = "Borough",
    y = "Poverty Rate (%)",
    caption = "\nSource: American Community Survey 5-Year Estimates, 2020-2024") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 3. Data and Variable Selection - EDIT TO VERIFY ACCURACY
Our primary dataset comes from the 2020–2024 American Community Survey (ACS) 
5-year estimates, which we access via the Census Bureau API at the census tract 
level across all five boroughs of New York City. The ACS 5-year estimates pool 
responses collected over a five-year period, making them the most reliable source 
of small-area data for geographies as granular as census tracts. Following Scally 
et al. (2021), we select variables across five thematic dimensions - economic 
conditions, educational attainment, housing, racial and ethnic composition, and 
mobility - to capture what makes neighborhoods genuinely distinct from one another, 
rather than relying on any single measure like income or poverty alone.

Many ACS variables are provided as raw counts, which reflect tract population size 
as much as neighborhood character. We therefore convert these to shares or rates. 
For example, expressing the number of residents below the poverty line as a 
proportion of all residents for whom poverty status is determined. Two variables, 
median household income and median gross rent, are kept as dollar values since they 
are already directly comparable across tracts. Mean commute time is calculated by 
dividing total travel time (B08013) by the number of workers aged 16 and over who 
commute outside the home (B08012), which is consistent with the Census Bureau's 
own approach to computing this measure.

We considered but decided to exclude two variables. First, the Gini coefficient (B19083), which measures income inequality within a tract, was dropped because it captures information already covered by our combination of median income and poverty rate. Second, age structure variables (B01001) were excluded because including multiple age-by-sex breakdown columns would add complexity without adding interpretive value; educational attainment serves as a cleaner proxy for the working-age population. Both decisions follow a simple guiding principle: where two variables tell essentially the same story, we keep the one that is easier to interpret in a policy context. 

We also wanted to consider a variable related to neighborhood safety as represented by crime. Ideally, we would have liked to use a crime rate by census tract variable. However, since that was not available, we decided to use NYPD Arrests data from the NYC Open Data Portal to calculate arrest rates by census tract. We used arrests instead of shootings because arrests also captures information on non-gun related violence and crimes, which may also contribute to the perception of safety in a neighborhood. To match arrests to census tracts, we also used the 2020 Census Centers of Population and identified arrests based on their proximity to the population center of the tract. We did this method as opposed to joining based on polygon geometry to more accurately capture the population's experience of crime, which is not strictly limited by census tract borders.

In addition to our primary 2020–2024 dataset, we also extract the 2018–2022 ACS 
5-year estimates for the same tracts. This allows us to compute four supplementary 
change variables in median income, median rent, poverty rate, and unemployment 
rate, that measure how conditions shifted between the two periods. One limitation 
worth noting is that because both ACS vintages include data collected between 2020 
and 2022, they share three overlapping years of data. This means the change 
variables capture a narrower window of genuine change than the four-year gap between 
the vintage endpoints might suggest. This is a structural feature of ACS 5-year 
estimates rather than a choice we made, and we flag it as a limitation of our 
approach.

Our 18 modeling variables are used as inputs to our dimensionality reduction and clustering models. Census tracts with missing values on any modelling variable are excluded from the analysis (why? I don't think we need to do that, let the method do it. We can also impute median for variables with not too many missing). Missing data arise most commonly in tracts that are entirely renter-occupied, which have no median home value, and in tracts with very few commuters, which have no reliable commute time estimate. The list of variables is included in @tbl-variables below.


```{r}
#| label: tbl-variables
#| tbl-cap: "Variable Selection and Justification"
#| echo: false

Dimension <- c("Economic", "Economic", "Economic", "Education", "Housing", "Housing", "Housing", "Housing", "Race/Ethnicity", "Race/Ethnicity", "Race/Ethnicity", "Race/Ethnicity", "Mobility", "Crime", "Change", "Change", "Change", "Change")
Variable <- c("Poverty rate", "Unemployment rate", "Median household income", "Bachelor's degree share", "Median gross rent", "Median home value", "Renter-occupied share", "Vacancy rate", "Share White", "Share Black", "Share Hispanic", "Share foreign-born", "Mean commute time", "Arrests per 1,000 residents", "Change in median income", "Change in median rent", "Change in poverty rate", "Change in unemployment")
Construction <- c("100*Pop. below poverty line / poverty universe total", "100*Unemployed civilians / civilian labor force", "Level (dollars)", "100*Bachelor's holders / pop. 25+", "Level (dollars)", "Level (dollars)",  "100*Renter units / total occupied units", "100*Vacant units / total housing units", "100*White alone / total population", "100*Black alone / total population", "100*Hispanic or Latino / total population", "100*Foreign-born / total population", "Aggregate travel time / commuting workers 16+", "1000*Total arrests/total population", "% change, 2022-2024 vintages", "% change, 2022-2024 vintages", "Percentage point change", "Percentage point change")
Justification <- c("Core deprivation indicator; directly named in proposal", "Captures labor market exclusion distinct from income poverty", "Primary measure of economic resources; retained as level as it is already comparable across tracts", "Captures human capital; strongly associated with neighborhood economic trajectory", "Directly named in proposal; key measure of housing cost", "Complements rent; distinguishes owner- vs. renter-dominated neighborhoods", "Tenure structure is a fundamental dimension of NYC neighborhood type", "Indicator of disinvestment; key differentiator between declining and stable neighborhoods", "Captures racial composition; connected to NYC's documented racial wealth gap", "As above", "As above, captures ethnicity composition", "NYC's immigrant geography is a major structural driver of neighborhood type", "Directly named in proposal; proxy for transit access and economic integration", "Captures information on neighborhood safety", "Captures change over time", "As above", "As above, percentage point change is appropriate for rates", "As above")

variable_table <- tibble(Dimension, Variable, Construction, Justification)

knitr::kable(variable_table)
```


```{r}
#| echo: false
#library(kableExtra)

#variable_table <- tribble(
#  ~Dimension,        ~Variable,                  ~Construction,                                          ~Justification,
#  "Economic",        "Poverty rate",              "Pop. below poverty line / poverty universe total",     "Core deprivation indicator; directly named in proposal",
  
#  "Economic",        "Unemployment rate",         "Unemployed civilians / civilian labor force",          "Captures labor market exclusion distinct from income poverty",
  
#  "Economic",        "Median household income",   "Level (dollars)",                                      "Primary measure of economic resources; retained as level as it is already comparable across tracts",
#  "Education",       "Bachelor's degree share",   "Bachelor's holders / pop. 25+",                       "Captures human capital; strongly associated with neighborhood # economic trajectory",
#  "Housing",         "Median gross rent",         "Level (dollars)",                                      "Directly named in proposal; key measure of housing cost burden",
#  "Housing",         "Median home value",         "Level (dollars)",                                      "Complements rent; distinguishes owner- vs. renter-dominated neighborhoods",
#  "Housing",         "Renter-occupied share",     "Renter units / total occupied units",                  "Tenure structure is a fundamental dimension of NYC neighborhood type",
#  "Housing",         "Vacancy rate",              "Vacant units / total housing units",                   "Indicator of disinvestment; key differentiator between declining and stable neighborhoods",
#  "Race/Ethnicity",  "Share White",               "White alone / total population",                       "Captures racial composition; connected to NYC's documented racial wealth gap",
#  "Race/Ethnicity",  "Share Black",               "Black alone / total population",                       "As above",
#  "Race/Ethnicity",  "Share Hispanic",            "Hispanic or Latino / total population (B03003)",       "Uses B03003 rather than B02001 as Hispanic identity is an ethnicity not captured by race alone",
#  "Race/Ethnicity",  "Share foreign-born",        "Foreign-born / total population",                      "NYC's immigrant geography is a major structural driver of neighborhood type",
#  "Mobility",        "Mean commute time",         "Aggregate travel time / commuting workers 16+",        "Directly named in proposal; proxy for transit access and economic integration",
#  "Supplementary",   "Change in median income",   "% change, 2022-2024 vintages",                         "Post-hoc cluster validation only; not a modelling input",
#  "Supplementary",   "Change in median rent",     "% change, 2022-2024 vintages",                         "As above",
#  "Supplementary",   "Change in poverty rate",    "Percentage point change",                              "As above; percentage point change is appropriate for rates",
#  "Supplementary",   "Change in unemployment",    "Percentage point change",                              "As above")

#variable_table |>
#  kbl(
#    col.names = c("Dimension", "Variable", "Construction", "Justification"),
#    align = c("l", "l", "l", "l")) |>
#   kable_styling(
#    bootstrap_options = c("striped", "hover", "condensed"),
#    full_width = TRUE,
#    font_size = 13) |>
#  column_spec(1, bold = TRUE, width = "10em") |>
#  column_spec(2, width = "12em") |>
#  column_spec(3, width = "14em") |>
#  column_spec(4, width = "20em") |>
#  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") |>
#  pack_rows("Modelling Variables", 1, 13) |>
#  pack_rows("Supplementary Variables — Used After Clustering To Validate Typologies, # Not As Modelling Inputs", 14, 17)
```

## 4. Methodology
Our analysis compares two methods of clustering census tracts. The first method uses Generalized Low Rank Models (GLRM) to reduce the complexity of our data, followed by Gaussian Mixture Modeling (GMM) for clustering neighborhood groups. The second method is simpler, using Principal Component Analysis (PCA) for dimension reduction and K-means clustering to compare for the final classification. Each step is described below.

### 4.1 Method 1

#### 4.1.1 Dimension Reduction with Generalized Low Rank Modeling


#### 4.1.2 Clustering with Gaussian Mixture Modeling

Our primary clustering method is Gaussian Mixture Modeling (GMM), 
implemented using `library(mclust)`.

Instead of measuring the distance from each tract to a centroid, GMM 
models each cluster as a multivariate normal distribution, meaning each cluster 
has its own center and its own shape, capturing the spread and correlation 
structure of the tracts within it. Tracts that are close to the center of a 
cluster's distribution are more likely to belong to it; tracts far from the 
center are less likely.

GMM also produces soft assignments rather than hard ones. Rather than 
assigning each tract definitively to a single cluster, GMM estimates the 
probability that each tract belongs to each cluster. These probabilities (called 
responsibilities) sum to one for each tract. A tract with a responsibility of 
0.95 for cluster 3 belongs there with high confidence; a tract with 
responsibilities of 0.52 and 0.48 across two clusters is genuinely ambiguous, 
and GMM makes that uncertainty visible rather than hiding it.

To select the number of clusters, we use the Bayesian Information Criterion 
(BIC). BIC works by balancing model fit against model complexity. A model with 
more clusters will always fit the data better, but BIC penalizes the additional 
parameters required, steering us toward solutions that are as simple as possible 
while still capturing meaningful structure in the data. In `library(mclust)`, 
the function automatically tries between 1 and 9 clusters across 14 different 
variance model specifications, and we select the model with the highest BIC 
value reported by the package. One small technical note: the standard statistical 
definition of BIC says lower values are better, but the library(mclust) package 
displays BIC with the opposite sign, meaning we look for the highest value on its 
output plots. Nevertheless, the end result is the same; we are still selecting the model that best balances fit and simplicity.

We do not automatically accept the number of clusters that scores best on BIC. If the statistically preferred solution produces neighborhood types that are too similar to one another to be meaningfully different in a policy context, we will consider a solution with fewer clusters and explain our reasoning for doing so.

### 4.2 Method 2

#### 4.2.1 Dimension Reduction with PCA
We begin with 18 variables across five thematic dimensions. Many of these 
variables are likely to be correlated with one another. For example, poverty 
rate and unemployment rate tend to move together, as do median income and 
bachelor's degree share. When highly correlated variables are fed directly into 
a clustering algorithm, they effectively get counted twice, overweighting the 
dimension of variation they share and distorting the resulting groups.

To address this, we use Principal Component Analysis (PCA) before clustering. 
PCA takes our 18 variables and transforms them into a smaller set of new 
variables called principal components. Each component is a weighted combination 
of the original variables, and the components are constructed so that they are 
uncorrelated with one another. The first component captures the largest share 
of overall variation in the data, the second captures the next largest share, 
and so on. Together, a small number of components can often represent most of 
the information in the original data with much less redundancy. Because our variables are measured in very different units (dollars, 
percentages, minutes) we normalize all variables before running PCA by 
subtracting the mean and dividing by the standard deviation so that no single variable dominates due to its scale alone.

To decide how many components to retain, we use a scree plot and inspect the cumulative proportion of variance explained by each additional component. As shown in Figure 2, the plot reveals a clear elbow around the fifth component, after which each additional component contributes diminishing returns. However, applying our threshold of retaining enough components to explain at least 80% of the total variation in the data, we retain nine components, which together explain 80.2% of the cumulative variance. The retained components are then passed forward as inputs to the clustering stage.


```{r}
#| echo: false
#| message: false
#| results: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(mclust)
library(tidyclust)
library(factoextra)
library(broom)
library(sf)
library(h2o)

## Load Data and Prepare Modeling Data

nyc_final <- st_read("data/nyc_final.shp")

# [I think we need to do some exploratory data analysis, here and in the writeup]
# [since it's part of our analysis process]

# drop geometry before modeling
nyc_data <- nyc_final |>
  st_drop_geometry()

# check for highly correlated predictors [do not assign this to anything]
# [i've added a recipe step that removes the ID vars for PCA but we need them to
# identify tracts at the end]
# [also need to figure out the best way to format this table if we use it]
nyc_data |>
  select(-GEOID, -NAME, -contyfp) |>
  cor(use = "complete.obs") |>
  round(2)

sum(!complete.cases(nyc_data))

# select only the modelling variables, dropping ID columns
nyc_vars <- nyc_data |>
  select(-GEOID, -NAME, -contyfp) |>
  mutate(across(everything(), as.numeric))

# impute missing values and replace infinite values with column means
nyc_vars_imputed <- nyc_vars |>
  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .))) |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>
  as.data.frame()

# remove any columns with zero variance
nyc_vars_imputed <- nyc_vars_imputed[, apply(nyc_vars_imputed, 2, sd) > 0]

# run PCA
pca_fit <- prcomp(nyc_vars_imputed, scale. = TRUE, center = TRUE)

# check cumulative variance explained
summary(pca_fit)

# scree plot
data.frame(
  component = 1:length(pca_fit$sdev),
  variance = (pca_fit$sdev^2 / sum(pca_fit$sdev^2)) * 100) |>
  ggplot(aes(x = component, y = variance)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Figure 2. Scree Plot - Variance Explained by Each Principal Component",
    subtitle = "\nThe elbow around component 5 indicates diminishing returns.\nNine components are retained to explain 80.2% of total variance.",
    x = "Principal Component",
    y = "Percent of Variance Explained") +
  theme_minimal()

# extract 9 components
pca_scores <- as.data.frame(pca_fit$x[, 1:9])
```

#### 4.2.2 K-Means Clustering
We use K-means clustering as our clustering step with this simpler method. K-means works by grouping census tracts into a fixed number of clusters, where each tract is 
assigned to the cluster whose center (called the centroid) it is closest to. 

The algorithm works as follows:
1. A set of K centroids are placed randomly in the data space
2. Each tract is assigned to its nearest centroid based on Euclidean distance
3. The centroids are recalculated as the mean of all tracts assigned to them
4. Steps 2 and 3 repeat until no tracts change cluster

Because the starting positions of the centroids are random, the results of 
K-means are not fully deterministic, running the algorithm twice can produce 
different solutions. We address this by using 100 random starts (`nstart = 100`) 
and retaining the best solution, which is standard practice.

K-means also requires us to specify the number of clusters K in advance. To 
choose K, we use three complementary methods. The first is the total 
within-cluster sum of squares, which measures how tightly packed each cluster 
is, we look for an "elbow" in the plot where adding more clusters stops 
producing meaningful improvements. The second is the average silhouette width, 
which measures how well each tract fits its assigned cluster compared to the 
nearest alternative cluster, with higher values indicating better-defined 
clusters. The third is the gap statistic, which compares the within-cluster 
dispersion to a reference distribution representing data with no real cluster 
structure, with higher values indicating more meaningful clusters. Where these 
three methods agree, we have greater confidence in our chosen K.

We note two important limitations of K-means. First, it assigns every tract to 
exactly one cluster with no measure of uncertainty; a tract that sits on the 
boundary between two neighborhood types is treated the same as one that sits 
squarely in the middle of a cluster. Second, as Milechin et al. (2024) found in 
a closely related study, the statistically optimal number of clusters and the 
most interpretable number are not always the same. As with GMM, we do not automatically accept the statistically optimal number of clusters.


## 5. Results
We first applied PCA to our 18 variables to reduce the complexity of the data before clustering. As shown in Figure 2 above, the first principal component explains approximately 27% of the total variance, with each subsequent component explaining progressively less. We retained nine components, which together explain 80.2% of the variance in the data, in line with the 80% threshold we set out in our methodology.

We then applied K-means clustering to the nine principal components. To decide on the number of clusters, we used two diagnostic plots: the elbow plot and the silhouette plot. The silhouette plot favored two clusters, while the elbow plot suggested that the gains from adding more clusters become small after around five. We chose five clusters because, while two clusters is statistically cleaner, it is not very useful from a policy perspective; it would essentially just divide NYC into deprived and non-deprived neighborhoods, which a single indicator like poverty rate could already tell us.

The five clusters each tell a different story about what kind of neighborhood they represent. Cluster 1 stands out because it combines very high poverty (46%) with relatively high incomes and rents, suggesting these are mixed neighborhoods with significant inequality within them (likely parts of Manhattan where very wealthy and very poor residents live in close proximity). Cluster 2 is the most affluent group, with the highest median income ($154k), the highest share of residents with a bachelor's degree (39%), and the lowest poverty rate (10%). Cluster 3 has moderate poverty and is predominantly Black (67%), capturing a set of neighbourhoods that face deprivation but in a different way to Cluster 4, which has the highest poverty rate (32%), the lowest median income ($47k), and is predominantly Hispanic (53%). Cluster 5 sits in the middle with moderate income, moderate rent, and racially mixed, likely reflecting outer borough neighborhoods like parts of Queens and Staten Island.

## 6. Policy Implications

## 7. Conclusion

## Bibliography
Center for New York City Affairs. "While the Top Three Percent of Wage Earners Get Richer, New York City's Low Wage Workers Risk Greater Poverty." 2024. https://www.centernyc.org/reports-briefs/while-the-top-three-percent-of-wage-earners-get-richer-new-york-citys-low-wage-workers-risk-greater-poverty

Milechin et al. (2024) ???? 

New York City Comptroller. "The Racial Wealth Gap in New York." December 6, 2023. https://comptroller.nyc.gov/reports/the-racial-wealth-gap-in-new-york/

Scally, Corianne Payton, Amanda Gold, Yipeng Su, Jorge Morales-Burnett, Eric Burnstein, and Patrick Spauster. "Reenvisioning Rural America: How to Invest in the Strengths and Potential of Rural Communities." 2021. https://reenvisioning-rural-america.urban.org/

U.S. Census Bureau. American Community Survey 5-Year Estimates, 2018–2022 and 2020–2024. Accessed via Census Bureau API using tidycensus in R.

Zewdie, Hiwot Y., Jamaica R. Robinson, Marc A. Adams, Anjum Hajat, Jana A. Hirsch, Brian E. Saelens, and Stephen J. Mooney. "A Tale of Many Neighborhoods: Latent Profile Analysis to Derive a National Neighborhood Typology for the US." Health & Place 86 (March 2024): 103209. https://pubmed.ncbi.nlm.nih.gov/38408408/